---
title: "Reproducible Research: Peer Assessment 1"
output: 
  html_document:
    keep_md: true
---

## Loading and preprocessing the data
```{r, echo=TRUE}
activityMonitoringData <- read.csv(file = "activity.csv", 
                                   na.strings = c("NA"), 
                                   colClasses = c(steps="numeric", date="Date", interval="integer"))
     
```

## What is mean total number of steps taken per day?
```{r, echo=TRUE}
totalStepsPerDay <- aggregate(x = activityMonitoringData$steps,
                              by = list(activityMonitoringData$date),
                              FUN = sum)
names(totalStepsPerDay) = c("date", "steps")
```

Here is a histogram of the total number of steps taken each day ignoring NAs:
```{r, echo=TRUE}
hist(totalStepsPerDay$steps, 
     breaks = 50,
     xlim = c(0, 25000),
     ylim = c(0, 12),
     main = "Histogram of Total Steps Per Day", 
     xlab = "Total Steps Taken Per Day")
```

## What is mean total number of steps taken per day?
```{r, echo=TRUE}
meanStepsPerDay <- mean(totalStepsPerDay$steps, na.rm = TRUE)
medianStepsPerDay <- median(totalStepsPerDay$steps, na.rm = TRUE)
```

Mean of the total number of steps taken per day: `r format(x = meanStepsPerDay,
                                                            big.mark = ",",
                                                            digits = 1,
                                                            small.mark = ".",
                                                            nsmall = 1,
                                                            na.rm = TRUE)`

Median of the total number of steps taken per day: `r format(x = medianStepsPerDay,
                                                            big.mark = ",",
                                                            digits = 1,
                                                            small.mark = ".",
                                                            nsmall = 1,
                                                            na.rm = TRUE)`


## What is the average daily activity pattern?
```{r, echo=TRUE}
avgStepsPerTimeInterval <- aggregate(x = activityMonitoringData$steps,
                                     by = list(activityMonitoringData$interval),
                                     FUN = "mean",
                                     na.rm = TRUE)
names(avgStepsPerTimeInterval) = c("interval", "avgSteps")

plot(avgStepsPerTimeInterval$interval,
     avgStepsPerTimeInterval$avgSteps,
     type = "l",
     main = "Average Daily Activity Pattern",
     xlab = "Time Interval",
     ylab = "Average Total Steps")

maxInterval <- as.character(avgStepsPerTimeInterval[which.max(avgStepsPerTimeInterval$avgSteps), "interval"])
```

5-minute interval which contains the maximum number of steps: `r maxInterval`

## Imputing missing values
Total number of missing values in the dataset: `r format(sum(is.na(activityMonitoringData$steps)), big.mark = ",")`

#### Replace all of the NA values with the average value for that particular 5-minute interval
```{r, echo=TRUE}

#### Create and fill a data.frame for all days with average data 
everyDayAverageData <- activityMonitoringData
everyDayAverageData$steps <- avgStepsPerTimeInterval$avgSteps

#### Substitute the average values into a new imputed data.frame
imputedActivityMonitoringData <- activityMonitoringData
imputedActivityMonitoringData$steps <- ifelse(is.na(imputedActivityMonitoringData$steps),
                                              everyDayAverageData$steps,
                                              imputedActivityMonitoringData$steps)

#### Now work with the imputed data
imputedTotalStepsPerDay <- aggregate(x = imputedActivityMonitoringData$steps,
                              by = list(imputedActivityMonitoringData$date),
                              FUN = sum)
names(imputedTotalStepsPerDay) = c("date", "steps")
```

Total number of missing values in the imputed dataset: `r format(sum(is.na(imputedActivityMonitoringData$steps)), big.mark = ",")`
Here is a histogram of the total number of steps taken each day including imputed data:
```{r, echo=TRUE}
hist(imputedTotalStepsPerDay$steps, 
     breaks = 50,
     xlim = c(0, 25000),
     ylim = c(0, 12),
     main = "Histogram of Total Steps Per Day With Imputed Data", 
     xlab = "Total Steps Taken Per Day")

imputedMeanStepsPerDay <- mean(imputedTotalStepsPerDay$steps, na.rm = TRUE)
imputedMedianStepsPerDay <- median(imputedTotalStepsPerDay$steps, na.rm = TRUE)
```

Mean of the total number of steps taken per day (with imputed data): `r format(x = imputedMeanStepsPerDay,
                                                                                   big.mark = ",",
                                                                                   digits = 1,
                                                                                   small.mark = ".",
                                                                                   nsmall = 1,
                                                                                   na.rm = TRUE)`

Median of the total number of steps taken per day (with imputed data): `r format(x = imputedMedianStepsPerDay,
                                                                                   big.mark = ",",
                                                                                   digits = 1,
                                                                                   small.mark = ".",
                                                                                   nsmall = 1,
                                                                                   na.rm = TRUE)`

Do these values differ from the estimates from the first part of the assignment? 
  - The mean is the same, but the median is slightly different.

What is the impact of imputing missing data on the estimates of the total daily number of steps?
  - The mean doesn't change since you are only adding in average values, but the daily totals will increase.

## Are there differences in activity patterns between weekdays and weekends?
```{r, echo=TRUE}
imputedActivityMonitoringData$weekDay <- factor(weekdays(imputedActivityMonitoringData$date))
imputedActivityMonitoringData$weekPart <- factor(ifelse(imputedActivityMonitoringData$weekDay == "Saturday" |
                                                             imputedActivityMonitoringData$weekDay == "Sunday",
                                                        "weekend",
                                                        "weekday"))

avgStepsPerIntervalWeekPart <- aggregate(x=imputedActivityMonitoringData$steps,
                                        by = list(imputedActivityMonitoringData$interval, imputedActivityMonitoringData$weekPart),
                                        FUN = "mean",
                                        na.rm = TRUE)
names(avgStepsPerIntervalWeekPart) = c("interval", "weekpart", "avgSteps")

library(lattice,)
xyplot(avgSteps ~ interval | weekpart, 
       data = avgStepsPerIntervalWeekPart, 
       type = "l",
       xlab = "Interval",
       ylab = "Number of steps",
       layout = c(1, 2))
```
